\speculake: Hiding Malware in Speculative Execution

### 2018-05-08 paper/discussion.tex

%TODO: This is a defense to Spectre, but not to Exspectre...
\subsubsection{SpecFence}
The \texttt{mfence} instruction requires that instructions earlier in the code
be retired before instructions after it. This ensures that memory is loaded and
stored in a particular order (in the real world). However, the \texttt{mfence}
instruction can be speculated past, meaning that caches can be affected
out-of-order even with this instruction.

One possible defense against \speculake is the introduction of a new
instruction: \texttt{specfence} which designates to the CPU a point that cannot 
be speculated past. This would require that loads/stores after the 
\texttt{specfence} instruction would be on hold until loads/stores from before
the \texttt{specfence} instruction have retired. Note the subtle difference from
\texttt{mfence}, we not only require that the loads/stores before a
\texttt{specfence} retire ahead of the loads/stores after the 
\texttt{specfence}, but more strongly: the loads/stores after a 
\texttt{specfence} do not begin to execute at all until the loads/stores before
the \texttt{specfence} retire.

This would possibly require a set of new API instructions in most basic
libraries allowing for functions that ensure no speculation (for security, no
data is leaked, no mystery instructions are run) and a separate set of functions
(as exist now) which allow speculation to occur (for performance, no sensitive
data is at risk).

We note that this instruction seems to already exist, albeit unintentionally.
The \texttt{rep} instruction repeats the execution of string instructions a set
number of times (set in a register). When set to repeat 90 or more times, these
instructions are not speculated past, requiring they complete and be retired
before looking to the next instruction. This appears to be due to the 
\texttt{rep} instruction adding additional $\mu$ ops per instruction, we found
that using the \texttt{rep} instruction caused $2n$ $\mu$ ops when the string
operation works on $n$ bytes~\cite{uop_article}. Thus enough repetitions would cause instructions
after the \texttt{rep} instruction to wait for the full repetition to be retired
before beginning their execution.

While this either new or repurposed instruction may provide some defense
against \speculake there are clear drawbacks: 
\begin{enumerate}
    \item Code using this instruction will be slower to execute by design. The
    purpose of this instruction is to deliberately slow the CPU and force it to
    wait. This will lead to a direct trade-off of speed vs. security.

    \item This will require a new set of APIs for basic library functions that
    load and store data. It would be detrimental to require all loads and stores
    never be speculated past, meaning there must be an option for choosing when
    certain loads and stores are not to be speculated past. This comes with all
    the usual drawbacks of modified APIs, un-updated systems, software engineers
    unaware of the new API, or of the significance, or when speculation is
    appropriate and when it is dangerous.
\end{enumerate}

\subsubsection{Malware Detection}
The world of computer security is often referred to as an arms race, hackers and
criminals racing to create and use exploits before security researchers and
software engineers can create defenses. As such we expect to see the same type
of race with this new class of malware.

@@

@@

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this model
of hidden computation applies more broadly. As of now there are no ways of 
debugging speculative computation meaning that the computation done there is effectively
isolated as with trusted execution environments. The security provided by the obfuscated 
key schedule makes \speculake useful in the domains of Digital Rights Management (DRM)
and whoitebox cryptography more generally. 
%% Approximate Computing

>>>>

\subsubsection{Operating System Defenses}
One defense to limit the capabilities of \speculake is to introduce privileged
instructions that grant the Operating System more control.

A necessary component of \speculake is the ability of a process to affect the
Branch History Buffer (BHB), which is used by all processes, allowing programs
to affect how separate programs choose to make indirect jumps.

To address this a new set of instructions allowing the OS to set and clear the
Branch History Buffer on context switches allows for processes to maintain a
useful branch history while disabling other processes from affecting their jump
decisions.

By design these new instructions grant the operating system control over the
state a program generates introducing possible attack vectors for malicious
Operating Systems to poison the Branch History Buffer or even allow \speculake
type attacks.

\subsubsection{Malware Detection}
\label{subsubsec:malware}
The world of computer security is often referred to as an arms race, hackers and
criminals racing to create and use exploits before security researchers and
software engineers can create defenses. As such we expect to see the same type
of race with this new class of malware.

@@

\subsubsection{Cache Poisoning}
As mentioned in Section~\ref{subsubsec:malware} a vital component of \speculake
is the ability to perform cache timings. Any mechanism that modifies the cache
can affect the success of \speculake-like attacks.

Drastic methods such as completely poisoning the cache will hinder this attack
while also forcing almost all loads to result in a cache miss, effectively
nullifying the point of cache.

However, more conservative methods such as randomly loading values into cache
can have a negative affect on \speculake-like attacks (causing the cache timing
to predict incorrect values) while not having too detrimental of an affect on
programs as a whole.

In response to this method of defense \speculake-like attacks can modify the
number of iterations and confidence thresholds necessary to determine a value.
As mentioned in Section~\ref{subsec:impl-turing} our experiments found that it
was only necessary to perform 50 iterations of our \texttt{target\_fn} in order
to gain 80\% confidence in the value returned. While only 50 iterations are
necessary, increasing this value to higher levels can lead to a significantly
higher confidence with very little performance detriment (especially in the
case of AES decryption).

\subsection{Resilience of \speculake}
Spectre works to use a cache side-channel to leak information from a process. As
such defenses to Spectre work to limit the information available to a
speculatively executing program. One such solution may be a new instruction such
as  \texttt{specfence}: an instruction which designates to the CPU a point that
cannot be speculated past. This would require that loads/stores after the
\texttt{specfence} instruction would be on hold until loads/stores from before
the \texttt{specfence} instruction have retired. The loads/stores after a
\texttt{specfence} do not begin to execute at all until the loads/stores before
the \texttt{specfence} retire.

However effective such an instruction may be against Spectre, it is not a
defense against \speculake. Such an instruction would be useful in restricting
the execution of a program to restrict the speculative capabilities of program,
thereby stopping information from being leaked. However, a malicious user
creating \speculake malware would simply choose to not use this instruction.

This demonstrates the versatility of speculative execution in general. While
defenses against Spectre are needed, a more general purpose solution to
speculative execution as a whole is necessary to address the wide range of its
capabilities.

% We note that this instruction seems to already exist, albeit unintentionally.
% The \texttt{rep} instruction repeats the execution of string instructions a set
% number of times (set in a register). When set to repeat 90 or more times, these
% instructions are not speculated past, requiring they complete and be retired
% before looking to the next instruction. This appears to be due to the 
% \texttt{rep} instruction adding additional $\mu$ ops per instruction, we found
% that using the \texttt{rep} instruction caused $2n$ $\mu$ ops when the string
% operation works on $n$ bytes~\cite{uop_article}. Thus enough repetitions would cause instructions
% after the \texttt{rep} instruction to wait for the full repetition to be retired
% before beginning their execution.

@@

\subsubsection{Alternative Applications}
We have proposed the \speculake model as a new malware threat, however, this model
of hidden computation applies more broadly. As of now there are no ways of 
debugging speculative computation meaning that the computation done there is effectively
isolated as with trusted execution environments. The security provided by the obfuscated 
key schedule makes \speculake useful in the domains of Digital Rights Management (DRM)
and whitebox cryptography more generally. 
%% Approximate Computing

### 2018-05-08 paper/implementation.tex

While the \speculake model acheives only moderate decryption speed, the 
critical computation of the decryption is done in the speculative world.

@@

To acheive a balance with speed the number of bits in each instruction is both 
fixed and minimized.  A variable length instruction would require that the 
\texttt{Flush+Reload} stage search the maximum number of bits on each round, and each 
aditional bit doubles the search space that the \texttt{Flush+Reload} stage must 
traverse.  So every bit shorter effectively doubles the throughput of the 
emulator, and there is effectively no advantage to allowing variable length 
instructions. Details of the SPASM instruction set can be found in 
Appendix~\ref{appendix:spasm}.

Using this model we have implemented multiple example programs that can be run
as encrypted binaries in an \speculake payload. A \textit{HelloWorld} program 
that prints to stdout. A \textit{FizzBuzz} program that demonstrates control 
flow and arithmetic operations while printing to stdout. And finally,
a \textit{ReverseShell} program that opens and connects to a socket before  
duplicating I/O file descriptors and executing a local shell.

@@

>>>>

While the \speculake model acheives only moderate decryption speed, the 
critical computation of the decryption is done in the speculative world. 
To further mask secure this decryption we make use of nested speculation 
to obfuscate the key schedule, discussed in section~\ref{subsec:nested-spec}.

@@

To acheive a balance with speed the number of bits in each instruction is both 
fixed and minimized.  A variable length instruction would require that the 
\texttt{Flush+Reload} stage search the maximum number of bits on each round, and each 
aditional bit doubles the search space that the \texttt{Flush+Reload} stage must 
traverse.  So every bit shorter effectively doubles the throughput of the emulator~\footnote{Until it doesn't}, 
and there is effectively no advantage to allowing variable length 
instructions.

In this model of computation there are no arguments as the speculative 
primitive provides a limited number of bits. Each opcode performs a 
specific operation on the registers in the state maintained by the emulator.
While other small instruction set architectures exist, they either allow 
exceptions to their instruction length, are too long even in reduced form, 
or did not have significant support to make them favorable for developers.

We have used 6 bits in the construction of this instruction set as our original
goal was to limit the length of each opcode as much as possible. Given the result 
demonstrated in figure~\ref{fig:spec_bandwidth} we have relaxed this goal a bit, as similar 
performance can be acheived within a range of bits and other application 
factors weigh on performance more than channel width.

Using this model we have implemented multiple example programs that can be run
as encrypted binaries in an \speculake payload. A \textit{HelloWorld} program 
that prints to stdout. A \textit{FizzBuzz} program that demonstrates control 
flow and arithmetic operations while printing to stdout. And finally,
a \textit{ReverseShell} program that opens and connects to a socket before  
duplicating I/O file descriptors and executing a local shell. Details of the 
SPASM instruction set can be found in Appendix~\ref{appendix:spasm}.

@@

We note that an analyst might attempt to locate the speculative entry point by
searching for AES-NI instructions in the payload program's dead code, ultimately
discovering the keys that it uses. However, we find it is possible to overcome
this by having the trigger program communicate the decryption key to the payload
program via the branch predictor.

To accomplish this, we use multiple speculative entry points, each that derive a
different decryption key before calling a common decryption routine. Since
the exact speculative entry point is determined by the trigger program, an
analyst cannot discover the decryption key directly from the payload program.

An analyst could still enumerate all potential entry points, testing each one
until they find one that correctly decrypts the ciphertext. In a 1~MB binary,
there are (at most) only 1~million entry points, providing just 20~bits of
security.

To increase security, we \emph{chain} entry points together. Rather than derive
the key from a single entry point, we have each potential entry point perform
another indirect jump that the CPU cannot immediately resolve, forcing it to
speculate while already executing speculatively. The predicted target of that jump
will also be determined by the trigger program. For example, if the trigger
program makes 30 training jumps, followed by 10 additional indirect jumps, and
the payload program performs the same 30 training jumps before a stall, the CPU
will predict the payload program will also perform the next 10 jumps. If each
jump has the potential to land in up to 1024 possible locations, this would
provide in total $1024^{10} = 2^{100}$ possible keys, or 100~bits of security.
We describe our implementation of this nested speculative execution concept in
Section~\ref{subsec:nested-spec}.

### 2018-05-08 paper/related-work.tex
Expanded SPASM VM and reorganized nested speculation sections

% Discovery & Followup
Given the numerous directions that were being investigated in reverse engineering 
Intel processors, multiple teams discovered the Spectre vulnerabilites concurrently including 
the teams associated with Kocher \textit{et al.}~\cite{spectre}, the Project Zero team 
at Google~\cite{project_zero},and various others~\cite{evtyushkin2018branchscope,maisuradze2018speculose}. 
These works propose multiple variants making use of banch prediction.  Follow 
up work on speculative execution side channels have also demonstrated
web-based vulnerabilites~\cite{genkin2018drive} as well as hardware isolation  
leaks in Intel's SGX~\cite{spectre_sgx}.

>>>>

% Discovery & Followup
Given the numerous directions that were being investigated in reverse engineering 
Intel processors, multiple teams discovered the Spectre vulnerabilites concurrently including 
the teams associated with Kocher \textit{et al.}~\cite{spectre}, the Project Zero team 
at Google~\cite{project_zero},and various others~\cite{evtyushkin2018branchscope,maisuradze2018speculose}. 
These works propose multiple variants making use of banch prediction.  Follow 
up work on speculative execution side channels have also demonstrated
web-based vulnerabilites~\cite{genkin2018drive} as well as hardware isolation  
leaks in Intel's SGX~\cite{spectre_sgx}. At the time of writing multiple sources
have claimed new speculative execution related exploits termed \textit{Spectre-NG}
have been reported to Intel and assigned CVE numbers. We have no details
on content, though intel has labeled four as critical and another four 
as intermediate security flaws~\cite{spectre_ng_gossip}.
